---
title: "POLS 207"
subtitle: "Problem Set 2"
author: "Villaseñor-Derbez, J.C."
output:
  pdf_document: 
    fig_caption: yes
    toc: no
    number_sections: no
header-includes:
  # - \usepackage{float}
  # - \floatplacement{figure}{H}
  - \usepackage{amsfonts}
# bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Problem 1: Analyzing Experimental Data

```{r}
# Load packages
suppressPackageStartupMessages({
  library(startR)
  library(Matching)
  library(ebal)
  library(here)
  library(tidyverse)
})

# Load data
olken_dat <- read.csv(here("ps2", "OlkenData.csv"))
```

## Estimate the average treatment effect in this new dataset, using the difference in means estimator

```{r}
olken_dat %>%
  group_by(treat_invite) %>%
  summarize(mean_est = mean(pct_missing, na.rm = T)) %>%
  ungroup() %>%
  mutate(treat_invite = ifelse(treat_invite == 0, "control", "treatment")) %>%
  spread(treat_invite, mean_est) %>%
  mutate(diff_means = treatment - control) %>%
  knitr::kable(
    booktabs = T,
    col.names = c(
      "Non-treatment",
      "Treatment",
      "Difference in means"),
    caption = "Difference in means estimate for % budget missing for projects where villagers were invited to public hearings."
  )
```

## Derive a simple estimator for the standard error of the above difference-in-means

Both $Y_1$ and $Y_0$ are independent from each other and describe two different random variables, which have means $\hat{Y_1}$ and $\hat{Y_0}$. Each of them has a variance around them, given by $\mathbb{V}(\hat{Y_1})$ and $\mathbb{V}(\hat{Y_0})$. The variance around the difference in means is then given by:

$$\mathbb{V}(\hat{Y_1} - \hat{Y_0})$$

Since the variance of a variable is given by the expectation of the squared difference between the value of a variable and the expectation of this value (*i.e.* $\mathbb{V}(X) = \mathbb{E}[(X - \mathbb{E}[X]) ^ 2]$), we can rewrite the above as:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{E}[((\bar{Y_1} - \bar{Y_0}) - \mathbb{E}[(\bar{Y_1} - \bar{Y_0})]) ^ 2]
$$

We can expand the terms in the expectation on the right hand side and obtain:


$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{E}[(\bar{Y_1} - \bar{Y_0} - \mathbb{E}[\bar{Y_0}] + \mathbb{E}[\bar{Y_1}]) ^ 2]
$$

Re-grouping and factorizing a $-1$ from the second term gives us:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{E}[((\bar{Y_1} - \mathbb{E}[\bar{Y_1}]) + (\bar{Y_0} - \mathbb{E}[\bar{Y_0}])) ^ 2]
$$

The first term ($(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])$) contains the deviations from the expectation for $\bar{Y_1}$, and the second term contains the deviations from the expectation for $\bar{Y_0}$.

We can expand the squared term and obtain:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{E}[(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])^2 + 2(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])(\bar{Y_0} - \mathbb{E}[\bar{Y_0}]) + (\bar{Y_0} - \mathbb{E}[\bar{Y_0}])^2]
$$

We can expand the outer-most expectation and obtain:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{E}[(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])^2 ]+ \mathbb{E}[(\bar{Y_0} - \mathbb{E}[\bar{Y_0}])^2] + 2(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])(\bar{Y_0} - \mathbb{E}[\bar{Y_0}])
$$


Per the deffinition of the variance, we would obtain that this is just:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{V}(\bar{Y_1}) + \mathbb{V}(\bar{Y_0}) + 2(\bar{Y_1} - \mathbb{E}[\bar{Y_1}])(\bar{Y_0} - \mathbb{E}[\bar{Y_0}])
$$
The last term on the right is just the covariance times a constant ($2$), which then gives us:

$$
\mathbb{V}(\bar{Y_1} - \bar{Y_0}) = \mathbb{V}(\bar{Y_1}) + \mathbb{V}(\bar{Y_0}) + 2\mathrm{cov}(\bar{Y_1}, \bar{Y_0})
$$

Since we assumed $\bar{Y_1}$ and $\bar{Y_0}$ to be independent from each other, we would expect $\mathrm{cov}(\bar{Y_1}, \bar{Y_0}) = 0$. Therefore, the standard error of the difference in means estimator is given by:

$$
SE = \sqrt{\frac{\sigma^2_{Y1}}{N_1} + \frac{\sigma^2_{Y2}}{N_2}}
$$

## Use the data to estimate the standard error you derived in [the previous exercise]

```{r}
# Get variances
sigma_y1 <- var(olken_dat$pct_missing[olken_dat$treat_invite == 1], na.rm = T)
sigma_y0 <- var(olken_dat$pct_missing[olken_dat$treat_invite == 0], na.rm = T)

# Get sample sizes
N1 <- sum(olken_dat$treat_invite == 1, na.rm = T)
N0 <- sum(olken_dat$treat_invite == 0, na.rm = T)

# Calculate Standard Errors
SE <- sqrt((sigma_y1 / N1) + (sigma_y0 / N0))
```

The standard error is `r SE`.


## Check the covariate balance in this dataset on all covariates (all variables that are not the treatment assignment or the outcome vectors).

```{r}
balance <- olken_dat %>%
  drop_na() %>%
  dplyr::select(-pct_missing) %>%
  MatchBalance(
    treat_invite ~ head_edu + mosques + pct_poor + total_budget,
    data = .,
    print.level = 0
  ) %>%
  baltest.collect(
    var.names = c(
      "head_edu",
      "mosques",
      "pct_poor",
      "total_budget"),
    after = F
  ) %>%
  as_tibble(rownames = "Covariate") %>%
  dplyr::select(-contains("qq"), -contains("pooled"))

knitr::kable(
  balance,
  booktabs = T,
  col.names = c(
    "Covariate",
    "Mean (Treatment)",
    "Mean (Control)",
    "Standardized difference",
    "Variance ratio",
    "T p-value",
    "KS p-value"
  ),
  caption = "Pre-matching balance of covariates."
)
```

```{r, fig.cap = "Pre-matching standardized difference in means of for covariates meassured for treadetd and untreated villages with projects."}
ggplot(data = balance, mapping = aes(x = Covariate, y = sdiff)) +
  geom_point(size = 4, shape = 21, fill = "steelblue", color = "black") +
  coord_flip() +
  ggtheme_plot() +
  geom_hline(yintercept = 0, linetype = "dashed")
```

\clearpage

## Now use regression to estimate the SATE (sample average treatment effect). Is this estimate different from the difference-in-means estimate? 

```{r, results = "asis"}

lm(pct_missing ~ treat_invite, data = olken_dat) %>% 
  stargazer::stargazer(.,
                       se = estimatr::starprep(.,
                                               se_type = "HC1"),
                       t.auto = T,
                       p.auto = T,
                       header = F,
                       title = "Estimate of the Sample Average Treatment Effect.",
                       single.row = T
                       )

```


## Using your answer for (b), conduct a t-test of the null hypothesis that SAT E = 0. You may use a normal approximation for the cutoff value.

```{r}
bar_y1 <- mean(olken_dat$pct_missing[olken_dat$treat_invite == 1], na.rm = T)
bar_y0 <- mean(olken_dat$pct_missing[olken_dat$treat_invite == 0], na.rm = T)

dif_means <- bar_y1 - bar_y0

t_score <- abs(dif_means / SE)

df <- N1 + N0 - 2 #N1 and N0 were calculated in b)

p_value <- 2 * pt(t_score, df = df, lower.tail = F) #calculate two-tailed

```

Student's t-test for unparied samples suggests that there are no differences in percent missing budget between treated and untreated groups ($t(565) = 0.7665;\;p = 0.4436$).

\clearpage

## Is the standard error of the OLS estimate different than the standard error of the differencein-means estimate? Why or why not?

The standard error of the OLS estimate is different to the standard error that I calculated in the difference in means. In the OLS estimation, I used heteroskedastic-robust standard errors. Furthermore, the SE's used in my derivation assumed that $\mathrm{cov}(\bar{Y_1}, \bar{Y_0}) = 0$, and ignored sampling with / without replacement issue.

## Re-estimate SAT E using three additional regression models

### One in which you include all pre-treatment covariates as additional linear predictors

```{r}
m1 <- lm(pct_missing ~ ., data = olken_dat)
```


### Another in which you include arbitrary functions of the covariates

```{r}
m2 <- lm(pct_missing ~ treat_invite + mosques * pct_poor + total_budget + I(total_budget^ 2),
         data = olken_dat)
```

###  A third in which you interact the treatment variable with a demeaned covariate ($X_i - \bar{X}$)

```{r}
olken_dat_demean <- olken_dat %>% 
  mutate_at(.vars = vars(head_edu, mosques, pct_poor, total_budget),
            .funs = function(x){x - mean(x, na.rm = T)})

m3 <- lm(pct_missing ~ treat_invite * total_budget,
         data = olken_dat_demean)

```


### Report the treatment effect estimates and their robust standard errors

```{r, results = "asis"}
models <- list(m1, m2, m3)

stargazer::stargazer(models,
                     se = estimatr::starprep(models,
                                             se_type = "HC1"),
                     t.auto = T,
                     p.auto = T,
                     header = F,
                     title = "Estimate of the Sample Average Treatment Effect.",
                     single.row = T
)
```


# Problem 2: Randomization Inference

##  is the sharp-null, and how does it compare to the null hypothesis we tested in the previous question?

The sharp-null states that the treatment effect is 0 for all units. This means that, unlike the previous example, the treated and untreated potential outcomes are the same for each observation.

## Why is the sharp null a convenient choice? I.e., what special property of the sharp null allows us to obtain a distribution of outcomes for a test statistic under this null?

The sharp null is a non-parametric test. This means that we de not make any assumptions about the distribution of the variable of interest. By the central limit theorem, we can then use a normal distribution (or approximate one with a t distribution).


## Write your own function in `R` that takes as arguments a vector of outcomes, Y , and the original treatment assignment vector, D, and produces:

- (i) a plot showing the distribution of the difference in means statistic under the sharp-null

- (ii) a vertical line representing the observed difference in means relative to this distribution

- (iii) a p-value for the difference in means statistic against the sharp-null.

\clearpage

```{r}
sharp_null_fxn <- 
  function(y = NULL, D = NULL, n_perms = 10000, seed = 42, ...){
    # Run checks
    ## Did the user specify all parameters?
    if(is.null(y))
    {stop("You did not specify a vector of outcomes.")
      }
    if(is.null(D)){
      stop("You did not specify a vector of treatments.")
      }
    n_obs <- length(y)
    # Are parameters the correct size?
    if(n_obs != length(D)){
      stop(paste0("y and D have different lengths(", n_obs," and ", length(D),")."))
      }
    # Is D the correct class?
    if(!is.logical(D)){
      stop("D must be a logical vector with TRUE or FALSE indicating treatment or control.")
      }
    # Is the suggested number of permutations large enough?
    if((n_perms / n_obs) < 10){
      warning("Your number of permutations might not be high enough")
      }
    
    # Get treated and control units
    treated <- y[D]
    not_treated <- y[!D]
    # Calculate true difference in means
    true_diff_in_means <- mean(treated, na.rm = T) - mean(not_treated, na.rm = T)
    
    # Set a random seed for reprodcibility
    set.seed(seed)
    # Create empty vector to save the omega-estimated
    omega_diff_in_means <- rep(NA, length = n_obs) #safer than numeric(length = n_obs)
    # iterare over n_perms
    for(omega in 1:n_perms){
      # Generate a random treatment assignment vector.
      # Sample without replacement to obtain same proportion of treated and untreated.
      D_omega <- sample(x = D, size = n_obs, replace = FALSE)
      # Get treated and not treated based on D_omega
      treated_omega <- y[D_omega]
      not_treated_omega <- y[!D_omega]
      # Calculate difference in means
      omega_diff_in_means[omega] <-
        mean(treated_omega, na.rm = T) - mean(not_treated_omega, na.rm = T)
    }
    # Plot the density of the iterated estimates
    plot(density(omega_diff_in_means), main = quo("Distribution of"~tau~"("~omega~")"))
    # Add a line with the true difference in means
    abline(v = true_diff_in_means, col = "red", lwd = 2, lty = "dashed")
    # Calculate probabilities
    p <- sum(omega_diff_in_means >= true_diff_in_means) / n_perms
}
```

## Apply your function to the data from the Olken experiment in the previous section. How do your results compare to the results under the t-test or regression? Which do you trust, and how do you interpret any differences?

```{r}
# Create vector of outcomes
y <- olken_dat$pct_missing

# Create treatment vector
D <- olken_dat$treat_invite == 1

# Run the sharp null test
sharp_null <- sharp_null_fxn(y = y, D = D)
```


With the previous run, `r sharp_null * 100`\% of the $\hat{\tau}(\omega)$ had a value equal to or greater than $\hat\tau_{ATE}$. In terms of our one-sided hypothesis, this means that only 24.25\% of the random treatment assignment vectors produced an effect size larger than or equal to our estimated effect. The t-test and the regression suggested that there was a 0.44 and 0.48 chance of randomly obtaining a value equal to or larger than the difference in means estimate. The sharp null tells us there is a greater probability (0.75) of this occurring at random. The interpretation is a bit different, given that the null hypothesis assumes no change on any unit.































